# Vectors

In this chapter, we're going to learn about vectors, one of the key building blocks of R programming. By the end of this chapter, you will know:

* What is a vector?
* How are vectors created?
* What are data types and how can I tell what sort of data I'm working with?
* What is metadata?
* How can I summarize a vector?

## What is a vector?

In R, every variable is a vector. Think of a set of contiguous cells in a spreadsheet. 

```{r eval=TRUE}
set.seed(1234)
e  <- rnorm(100)
X1 <- 1:10
```

Here, `e` is a vector with `N` values. `X1` is the sequence of integers from 1 through 10. 

Vectors can grow and shrink automatically. No need to move cells around on a sheet. No need to copy formulas or change named ranges.

### Vector properties

* Every element in a vector must be the same type. 
    * R will change data types if they are not!
    * Different types are possible by using a list or a data frame (later)
* Vectors have one dimension
    * Higher dimensions are possible via matrices and arrays
* Possible to add metadata (like names) via attributes

### Vector construction

Vectors are constructed in one of several ways:

* Return from a function or operation
    * `seq`, `rep`, `sample`, `rnorm`,  etc.
* Concatenation
* Growth by assignment

### `seq`

`seq` is used often to generate a sequence of values. The colon operator `:` is a shortcut for a sequence of integers.

```{r }
pies = seq(from = 0, by = pi, length.out = 5)
i <- 1:5
year = 2000:2004
```

```{r echo=FALSE}
library(rblocks)
i <- 1:5
block = make_block(i, type = 'vector')
block
```

### `rep`

The `rep` function will replicate its input 
```{r }
i = rep(pi, 100)
head(i)
```

### Concatenation

The `c()` function will concatenate values.

```{r results='hide'}
i <- c(1, 2, 3, 4, 5)
j <- c(6, 7, 8, 9, 10)
k <- c(i, j)
l <- c(1:5, 6:10)
```

```{r echo=FALSE}
i = c(1, 2, 3, 4, 5)
j = c(6, 7, 8, 9, 10)
k = c(i, j)
block_i = make_block(i, type = 'vector')
block_j = make_block(j, type = 'vector')
block_i[1:5] = "red"
block_j[1:5] = "blue"
block_k = make_block(c(block_i,block_j), type='vector')
block_k[1:5] = "red"
block_k[6:10] <- "blue"

block_i
block_j


block_k
```

### Growth by assignment

Assigning a value beyond a vectors limits will automatically grow the vector. Interim values are assigned `NA`.

```{r }
i <- 1:10
i[30] = pi
i
```

### Vector access - by index

Vectors may be accessed by their numeric indices. Remember, ':' is shorthand to generate a sequence.

```{r }
set.seed(1234)
e <- rnorm(100)
e[1]
e[1:4]
e[c(1,3)]
```

### Vector access - logical access

Vectors may be accessed logically. This may be done by passing in a logical vector, or a logical expression.

```{r }
i = 5:9
i[c(TRUE, FALSE, FALSE, FALSE, TRUE)]
i[i > 7]
b = i > 7
b
i[b]
```

### `which`

The `which` function returns indices that match a logical expression.

```{r }
i <- 11:20
which(i > 12)
i[which(i > 12)]
```

### `sample`

The `sample` function will generate a random sample. Great to use for randomizing a vector.

```{r }
months <- c("January", "February", "March", "April"
            , "May", "June", "July", "August"
            , "September", "October", "November", "December")

set.seed(1234)
mixedMonths <- sample(months)
head(mixedMonths)
```

Get lots of months with the `size` parameter:

```{r }
set.seed(1234)
lotsOfMonths <- sample(months, size = 100, replace = TRUE)
head(lotsOfMonths)
```

### `sample` II

Sample may also be used within the indexing of the vector itself:

```{r }
set.seed(1234)
moreMonths <- months[sample(1:12, replace=TRUE, size=100)]
head(moreMonths)

# Cleaner with sample.int
set.seed(1234)
evenMoreMonths <- months[sample.int(length(months), size=100, replace=TRUE)]
head(evenMoreMonths)
```

### `order`

The function `order` will return the indices of the vector in order.

```{r }
set.seed(1234)
x <- sample(1:10)
x
order(x)
x[order(x)]
```

### Vector arithmetic

Vectors may be used in arithmetic operations. 
  
```{r eval=FALSE}
B0 <- 5
B1 <- 1.5

set.seed(1234)

e <- rnorm(N, mean = 0, sd = 1)
X1 <- rep(seq(1,10),10)

Y <- B0 + B1 * X1 + e
```

Y is now a vector with length equal to the longest vector used in the calculation.

Question: B0 and B1 are vectors of length 1. 

X1 and e are vectors of length 100. 

How are they combined?


### Recycling

R will "recycle" vectors until there are enough elements to perform an operation. Everything gets as "long" as the longest vector in the operation. For scalar operations on a vector this doesn't involve any drama. Try the following code:

```{r size='tiny'}
vector1 = 1:10
vector2 = 1:5
scalar = 3

print(vector1 + scalar)
print(vector2 + scalar)
print(vector1 + vector2)
```

### Set theory - Part I

The `%in%` operator will return a logical vector indicating whether or not an element of the first set is contained in the second set.

```{r }
x <- 1:10
y <- 5:15
x %in% y
```

### Set theory - Part II

* `union`
* `intersect`
* `setdiff`
* `setequal`
* `is.element`

```{r eval = FALSE}
?union
```

```{r }
x <- 1900:1910
y <- 1905:1915
intersect(x, y)
setdiff(x, y)
setequal(x, y)
is.element(1941, y)
```

### Summarization

Loads of functions take vector input and return scalar output. Translation of a large sest of numbers into a few, informative values is one of the cornerstones of statistics.

```{r eval=FALSE}
x = 1:50
sum(x)
mean(x)
max(x)
length(x)
var(x)
```

### Vectors 

Vectors are like atoms. If you understand vectors- how to create them, how to manipulate them, how to access the elements, you're well on your way to grasping how to handle other objects in R. 

Vectors may combine to form molecules, but fundamentally, _everything_ in R is a vector.

## From data



* Data types
* From vectors to matrices and lists

### Data types

* logical
* integer
* double
* character

### What is it?

```{r}
x <- 6
y <- 6L
z <- TRUE
typeof(x)
typeof(y)
typeof(z)
is.logical(x)
is.double(x)
```

### Data conversion

Most conversion is implicit. For explicit conversion, use the `as.*` functions. 

Implicit conversion alters everything to the most complex form of data present as follows:

logical -> integer -> double -> character

Explicit conversion usually implies truncation and loss of information.

```{r }
# Implicit conversion
w <- TRUE
x <- 4L
y <- 5.8
z <- w + x + y
typeof(z)

# Explicit conversion. Note loss of data.
as.integer(z)
```

### Class

A class is an extension of the basic data types. We'll see many examples of these. The class of a basic type will be equal to its type apart from 'double', whose class is 'numeric' for reasons I don't pretend to understand.

```{r }
class(TRUE)
class(pi)
class(4L)
```
The type and class of a vector is returned as a scalar. Remember a vector is a set of elements which all have the same type.
```{r }
class(1:4)
```

### Mode

There is also a function called 'mode' which looks tempting. Ignore it.

### Dates and times

Dates in R can be tricky. Two basic classes: `Date` and `POSIXt`. The `Date` class does not get more granular than days. The `POSIXt` class can handle seconds, milliseconds, etc.

My recommendation is to stick with the "Date" class. Introducing times means introducing time zones and possibility for confusion or error. Actuaries rarely need to measure things in minutes.

```{r }
x <- as.Date('2010-01-01')
class(x)
typeof(x)
```

### More on dates

The default behavior for dates is that they don't follow US conventions.

Don't do this:
```{r error=TRUE}
x <- as.Date('06-30-2010')
```

But this is just fine:
```{r }
x <- as.Date('30-06-2010')
```

If you want to preserve your sanity, stick with year, month, day.
```{r }
x <- as.Date('2010-06-30')
```

### What day is it?

To get the date and time of the computer, use the either `Sys.Date()` or `Sys.time()`. Note that `Sys.time()` will return both the day AND the time as a POSIXct object.

```{r }
x <- Sys.Date()
y <- Sys.time()
```

### More reading on dates

Worth reading the documentation about dates. Measuring time periods is a common task for actuaries. It's easy to make huge mistakes by getting dates wrong.

The `lubridate` package has some nice convenience functions for setting month and day and reasoning about time periods. It also enables you to deal with time zones, leap days and leap seconds. Probably more than you need.

`mondate` was written by an actuary and supports (among other things) handling time periods in terms of months.

* [Date class](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Dates.html)
* [lubridate](http://www.jstatsoft.org/v40/i03/paper)
* [Ripley and Hornik](http://www.r-project.org/doc/Rnews/Rnews_2001-2.pdf)
* [mondate](https://code.google.com/p/mondate/)

### Factors

Another gotcha. Factors were necessary many years ago when data collection and storage were expensive. A factor is a mapping of a character string to an integer. Particularly when importing data, R often wants to convert character values into a factor. You will often want to convert a factor into a string.

```{r }
myColors <- c("Red", "Blue", "Green", "Red", "Blue", "Red")
myFactor <- factor(myColors)
typeof(myFactor)
class(myFactor)
is.character(myFactor)
is.character(myColors)
```

### Altering factors

```{r }
# This probably won't give you what you expect
myOtherFactor <- c(myFactor, "Orange")
myOtherFactor

# And this will give you an error
myFactor[length(myFactor)+1] <- "Orange"

# Must do things in two steps
myOtherFactor <- factor(c(levels(myFactor), "Orange"))
myOtherFactor[length(myOtherFactor)+1] <- "Orange"
```

### Avoid factors

Now that you know what they are, you can spend the next few months avoiding factors. When R was created, there were compelling reasons to include factors and they still have some utility. More often than not, though, they're a confusing hindrance.

If characters aren't behaving the way you expect them to, check the variables with `is.factor`. Convert them with `as.character` and you'll be back on the road to happiness.

### Questions

* Create a logical, integer, double and character variable.
* Can you create a vector with both logical and character values?
* What happens when you try to add a logical to an integer? An integer to a double?

### Answers
```{r }
myLogical <- TRUE
myInteger <- 1:4
myDouble <- 3.14
myCharacter <- "Hello!"

y <- myLogical + myInteger
typeof(y)
y <- myInteger + myDouble
typeof(y)
```

### From vectors to matrices and lists

A matrix is a vector with higher dimensions.

A list has both higher dimensions, but also different data types.

### A matrix

Two ways to construct:

1. Use the `matrix` function.
2. Change the dimensions of a `vector`.

```{r }
myVector <- 1:100
myMatrix <- matrix(myVector, nrow=10, ncol=10)

myOtherMatrix <- myVector
dim(myOtherMatrix) <- c(10,10)

identical(myMatrix, myOtherMatrix)
```

### 

```{r }
myMatrix <- matrix(nrow=10, ncol=10)
```

```{r echo=FALSE}
library(rblocks)
block_grid(10, 10, type="matrix")
```

### 

```{r }
dim(myMatrix) <- c(25, 4)
```

```{r echo=FALSE}
block_grid(25, 4, type="matrix")
```

### Matrix metadata

Possible to add metadata. This is typically a name for the columns or rows.

```{r }
myMatrix <- matrix(nrow=10, ncol=10, data = sample(1:100))
colnames(myMatrix) <- letters[1:10]
head(myMatrix, 3)
rownames(myMatrix) <- tail(letters, 10)
head(myMatrix, 3)
```

### Data access for a matrix

Matrix access is similar to vector, but with additional dimensions. For two-dimensional matrices, the order is row first, then column. 

```{r }
myMatrix[2, ]
myMatrix[, 2]
```

### Data access continued

Single index will return values by indexing along only one dimension.

```{r }
myMatrix[2]
myMatrix[22]
```

### Matrix summary

```{r }
sum(myMatrix)
colSums(myMatrix)
rowSums(myMatrix)
colMeans(myMatrix)
```

### More than two dimensions

Like more than two dimensions? Shine on you crazy diamond.


## Exercise

Create a vector of length 10, with years starting from 1980.

Create a vector with values from 1972 to 2012 in increments of four (1972, 1976, 1980, etc.)

Construct the following vectors (feel free to use the `VectorQuestion.R` script):
```{r }
FirstName <- c("Richard", "James", "Ronald", "Ronald"
              , "George", "William", "William", "George"
              , "George", "Barack", "Barack")
LastName <- c("Nixon", "Carter", "Reagan", "Reagan"
              , "Bush", "Clinton", "Clinton", "Bush"
              , "Bush", "Obama", "Obama")
ElectionYear <- seq(1972, 2012, 4)
```

* List the last names in alphabetical order
* List the years in order by first name.
* Create a vector of years when someone named "George" was elected.
* How many Georges were elected before 1996?
* Generate a random sample of 100 presidents.

## Answers

```{r }
LastName[order(LastName)]
ElectionYear[order(FirstName)]
ElectionYear[FirstName == 'George']
myLogical <- (FirstName == 'George') & (ElectionYear < 1996)
length(which(myLogical))
sum(myLogical)

sample(LastName, 100, replace = TRUE)
```
