# First Steps {-}

The first part of this book will get you started using R. Unless you've been using R comfortably for several months, resist the tempation to skip through this quickly. We've found that one of the biggest hurdles to getting users comfortable with R is getting the software installed and running with a minimum of hassle. The next step is getting folks reconciled to the idea that R is very different from Excel. You can like that, or you can hate that, but you can't change that. Hopefully, by the end of this section, you'll be open to the idea that R can accomplish some tasks and it's something you can add to your tool kit.

```{r echo=FALSE, results='hide'}
library(pander)
```

# Setup

By the end of this chapter, you will have done the following:

* Install R
* Install RStudio
* Install the `raw` package

## Operating systems

Although R was developed primarily on Unix-based operating systems, it may be used on many different platforms. As I write these words, there are three major systems in use: Windows, Mac OS and Linux. I've used R and RStudio on all three and the experience is pretty much the same. This is one of the fantastic features of the software. It's meant to be as widely used and portable as possible to maximize its use. 

There are one or two operating system quirks, but in general I won't need to OS differences again, apart from one preliminary note. When refrring to keystroke combinationsn, I will only refer to the CTRL key. Mac OS users will understand that this key is CMD on their keyboards. 

If you're curious, the version of R and system architecture being used to write this book are noted below:

```{r echo = FALSE}
sessionInfo()$platform
```

### Installing R

In each case, what you'll do is download a file from the internet and then follow the standard process you go through to install software on whichever system you're using. For the most part, installation is quick and painless, but there may be limitations placed on you by your IT department. I have a few suggestions which I hope can help overcome any difficulties you might experience.

The first place to look for installation is [cran.r-project.org]. From there, you will see links to downloads for Windows, Mac and Linux. Clicking on the appropriate link will take you to the page that's relevant for your operating system. You may see lots of bizarre, arcane language around binaries, source and tarballs. If those words (in this context) mean nothing to you, don't panic. Some folk like to build their own version of R directly from the source code. If you're reading these instructions, you're probably not one of those people. 

I recommend getting familiar with the CRAN website and reading the documentation there. If you get totally lost, try the links below which should take you directly to the download site for Windows and Mac. (If you're running Linux, I can't imagine you need my help.)

* [Windows install](http://cran.revolutionanalytics.com/bin/windows/base/)
* [Mac install](http://cran.revolutionanalytics.com/bin/macosx/)

https://cran.r-project.org/bin/windows/base/

It's possible that you'll be asked to identify a "mirror". R is hosted on a number of servers throughout the world. It's all the same R, but distributing it in this way helps to minimized load on servers which host the files. 

### Bitness

32 vs 64

### Installing R Studio

Installing R is most of the battle. Depending on the sort of person you are, it may even be all of the battle (see the following section on environments). R comes with a fairly spartan user interface, which is sufficient to get work done. However, most folk find that they enjoy using an Integrated Development Environment (IDE). This allows one to work on several source files at the same time, read help, observe console output, see what variables are loaded in memory, etc. There are a few options, but I've not yet found anything better than RStudio.

RStudio's main website may be found at [www.rstudio.com](www.rstudio.com). At the time of writing, the download page may be found at [http://www.rstudio.com/products/rstudio/download/](http://www.rstudio.com/products/rstudio/download/). Here you will find links to specific systems. The browser will even attempt to detect what operating system you're using and suggest a link for you. Cool, huh?

### IT

I don't think I've ever met anyone who's made it through a white-collar existence without at least one or two frustrating exchanges with a corporate IT department. If you work for a large- or even small- company, you likely have a staff of folks who keep the network running and handle software requests from every user in the company. To ensure that your company's network is free from malicious attack or well-intentioned, but careless or imperfect users, most computers have sensitive areas restricted. This means that if you want to install software, you need an administrator to do it for you.

What this also means is that your IT department might not be as delighted as you are to install open-source software on the company laptop. This might be a problem that they're not inclined to solve for you and you may find your interaction with IT folks to a bit frustrating and they may seem as though they're not at all helpful.

The first thing to bear in mind is that, despite any appearance to the contrary, your IT staff is there to help you. Moreover, they're people. They have families who love them, possibly small children who think their moms and dads are awesome, pets who miss them and lives outside of work. They have to deal with ridiculous hours to accommodate you and they get far more complaints than they do praise. Be nice to them and you may be surprised how supportive they can be.

With that understood, there are several situations you may find yourself in.:

* You have- or can talk your way into- admin rights to your computer

Lucky you. Also lucky me as this is the happy situation that I enjoy. How do you handle this situation? Don't blow it! Be careful what you download, don't greedily consume bandwidth, server space or any of your company's other scarce resources and be VERY NICE to your IT staff. Acknowledge that you're grateful to have been given such trust and pledge not to do anything to have it removed.

* You don't have admin rights to your computer

What to do? Request to be given admin rights. Explain why, in detail and don't be evasive or vague. Trust and mutual respect help. Talk to other folks in your department and get them on board. Present a strong business case for why use of this software will permit you and your department to work more efficiently. Show them this book and underline the parts where I tell folk to be nice and respectful towards IT staff.

* IT won't give you admin rights to your computer

In this case, you may ask them to install it for you. Pool your resources. Talk to other actuaries and analysts in your company. Talk to your boss.

* IT won't install the software.

Solution? Install the software to a memory stick. Yes, it is often (but not always!) possible to do this. This is is obviously not a preferred option, but it will get you up and running and enable you to attend the workshop.

* Memory sticks are locked down

In this case, your IT department really wants you all to be running terminals. OK. Suggest an install of RStudio Server. This enables R to run on a server with controlled user access. This is quite a lot more work for your IT staff and you'll need to make a strong use case for it. If you're a large organization which has a predictive modelling or analytics area, they'll likely want this software. This won't allow you to use R remotely, so getting the most out of the workshop will be tough. However, there's still one more option.

* The nuclear option

Your IT staff won't run R on a server, won't give you a laptop with R installed. They're really against this software. I'd like to advise you to get another job, but that's defeatist. This is where we reach the nuclear option, which is to use your own computer. This will drive folks at your company nuts. Now you're transferring data from a secure machine to one which you use for personal e-mail, Facebook, sports, personal finance and other activities that we needn't dwell on here. This is an absolute last resort and the overheard of moving stuff from one device to another will obviate most of the efficiency gains that open source software will provide. Here's how to make it work: produce work that is ONLY POSSIBLE using R, or Python or any of the tools which we will discuss. Show a killer visual and then patiently explain to your boss why it can't be done in Excel and why you can't share it with other departments and why it can't be done every quarter. This is a tall order, but it just might get someone's attention.